---
title: "Normal Approximation and Sample Size Calculation"
format: "tlda-labs/tlda-revealjs"
# knitr caching off for this deck: the build box fails writing the cache (2026-10-06)
execute:
  cache: false
---



# Review

{{< include shared/shared-code.qmd >}}
{{< include ga-turnout.qmd >}}

## Where We Left Off: Exact and Guaranteed

```{r}
#| label: intcomp-data
#| include: false
library(ggplot2)

# Plot helper only. All statistical work lives in visible intcomp-calc.
# No legend is drawn on the plot: the slide text beside each figure carries
# the pink/teal key in matching colors.
intcomp.plot = function(estimator, state=3) {
  stopifnot(estimator %in% c("binomial", "hypergeometric"), state %in% 1:3)
  if (estimator == "binomial") {
    samples = intcomp.samples.binom
    hw.exact = intcomp.hw.exact.binom
    hw.cheb = intcomp.hw.cheb.binom
    pmf = intcomp.pmf.binom
  } else {
    samples = intcomp.samples.hyper
    hw.exact = intcomp.hw.exact.hyper
    hw.cheb = intcomp.hw.cheb.hyper
    pmf = intcomp.pmf.hyper
  }
  # bootstrap = binomial plug-in at each sample's own proportion; it doesn't know m
  hw.boot = sapply(samples[1:100], \(p) width_binom(p, intcomp.n)/2)
  hws = list(hw.exact, hw.cheb, hw.boot)
  colors = c(pink, teal, 'black')
  offsets = c(0, .04, -.04)
  # all 100 sample means carry the current technique's interval; the highlighted
  # sample keeps every technique so far: exact, Chebyshev above it, bootstrap below.
  current = with_annotations(samples, height=1, dots=100, highlight.dots=1,
                 interval.width=2*hws[[state]], vline.alpha=NA, seed=default.seed)
  current[[3]]$data$color[1] = colors[state]
  current[[3]]$data$y[1] = current[[3]]$data$y[1] + offsets[state]
  hero = lapply(seq_len(state - 1), function(k) {
    g = with_annotations(samples, height=1, dots=100, highlight.dots=1,
          interval.width=2*hws[[k]], vline.alpha=NA, seed=default.seed,
          unhighlighted.dot.alpha=0)
    g[[3]]$data$color[1] = colors[k]
    g[[3]]$data$y[1] = g[[3]]$data$y[1] + offsets[k]
    g
  })
  sampling.dist = geom_col(aes(x=(0:intcomp.n)/intcomp.n, y=pmf/max(pmf)),
                           alpha=.1, linewidth=.1, fill='black', color='black')
  estimand = geom_vline(xintercept=intcomp.theta, color='green', linewidth=3, alpha=.8)
  limits = range(c(intcomp.samples.binom[1:100]-intcomp.hw.cheb.binom,
                   intcomp.samples.binom[1:100]+intcomp.hw.cheb.binom,
                   intcomp.samples.hyper[1:100]-intcomp.hw.cheb.hyper,
                   intcomp.samples.hyper[1:100]+intcomp.hw.cheb.hyper))
  ggplot() + sampling.dist + estimand + current + hero +
    coord_cartesian(xlim=limits+c(-.005,.005), ylim=c(0,1.01)) +
    scale_y_continuous(breaks=NULL) +
    labs(x=NULL, y=NULL) +
    simplified.theme
}
```

```{r}
#| label: intcomp-calc
# Model parameters (explicit; change them here). Hypergeometric population
# m chosen so its proportion matches theta (m*theta must be an integer
# count); FPC = (m-n)/(m-1) enters its actual variance.
intcomp.n = 625
intcomp.theta = 0.7
intcomp.m = 2000
intcomp.samples.binom = with_seed(
  rbinom(10000, intcomp.n, intcomp.theta)/intcomp.n)
intcomp.samples.hyper = with_seed(
  rhyper(10000, intcomp.m*intcomp.theta,
         intcomp.m*(1-intcomp.theta), intcomp.n)/intcomp.n)
# Actual PMFs and true SDs.
intcomp.pmf.binom = dbinom(0:intcomp.n, intcomp.n, intcomp.theta)
intcomp.pmf.hyper = dhyper(0:intcomp.n, intcomp.m*intcomp.theta,
                           intcomp.m*(1-intcomp.theta), intcomp.n)
intcomp.sigma.binom = sqrt(intcomp.theta*(1-intcomp.theta)/intcomp.n)
intcomp.sigma.hyper =
  sqrt(intcomp.theta*(1-intcomp.theta)/intcomp.n *
       (intcomp.m-intcomp.n)/(intcomp.m-1))
# Exact central-95 half-widths from SHARED width_binom/width_hyper only.
intcomp.hw.exact.binom = width_binom(intcomp.theta, intcomp.n)/2
intcomp.hw.exact.hyper = width_hyper(intcomp.theta, intcomp.n, intcomp.m)/2
# Chebyshev 95% half-widths: 1/t^2 = .05 gives t = sqrt(20); each estimator
# uses its ACTUAL true model sigma, never a sample plug-in.
intcomp.hw.cheb.binom = sqrt(20)*intcomp.sigma.binom
intcomp.hw.cheb.hyper = sqrt(20)*intcomp.sigma.hyper
# Measured integer-support coverage of those widths (measurement only: sums
# the pmf over counts inside each interval; finds no widths).
intcomp.measured.coverage = function(pmf, hw) {
  k = 0:intcomp.n
  inside = k >= ceiling(intcomp.n*(intcomp.theta - hw)) &
           k <= floor(intcomp.n*(intcomp.theta + hw))
  sum(pmf[inside])
}
intcomp.cov.exact.binom = intcomp.measured.coverage(intcomp.pmf.binom,
                                                    intcomp.hw.exact.binom)
intcomp.cov.exact.hyper = intcomp.measured.coverage(intcomp.pmf.hyper,
                                                    intcomp.hw.exact.hyper)
intcomp.cov.cheb.binom = intcomp.measured.coverage(intcomp.pmf.binom,
                                                   intcomp.hw.cheb.binom)
intcomp.cov.cheb.hyper = intcomp.measured.coverage(intcomp.pmf.hyper,
                                                   intcomp.hw.cheb.hyper)
```

::::: columns
:::: column
Sampling with replacement

:::: {.r-stack}
::: {.fragment .fade-out fragment-index="0"}
```{r}
#| label: fig-intcomp-binom-0
#| echo: false
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-alt: "Binomial sampling distribution with one hundred sample means; every sample mean carries its interval: exact in the first state, Chebyshev in the second, bootstrap in the third, with one sample highlighted. Green line at the population proportion. State 1 of 3."
intcomp.plot("binomial", 1)
```
:::
::: {.fragment .current-visible fragment-index="0"}
```{r}
#| label: fig-intcomp-binom-1
#| echo: false
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-alt: "Binomial sampling distribution with one hundred sample means; every sample mean carries its interval: exact in the first state, Chebyshev in the second, bootstrap in the third, with one sample highlighted. Green line at the population proportion. State 2 of 3."
intcomp.plot("binomial", 2)
```
:::
::: {.fragment fragment-index="1"}
```{r}
#| label: fig-intcomp-binom-2
#| echo: false
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-alt: "Binomial sampling distribution with one hundred sample means; every sample mean carries its interval: exact in the first state, Chebyshev in the second, bootstrap in the third, with one sample highlighted. Green line at the population proportion. State 3 of 3."
intcomp.plot("binomial", 3)
```
:::
::::
::::
:::: column
Sampling without replacement

:::: {.r-stack}
::: {.fragment .fade-out fragment-index="0"}
```{r}
#| label: fig-intcomp-hyper-0
#| echo: false
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-alt: "Hypergeometric sampling distribution with one hundred sample means; every sample mean carries its interval: exact in the first state, Chebyshev in the second, bootstrap in the third, with one sample highlighted. Green line at the population proportion. State 1 of 3."
intcomp.plot("hypergeometric", 1)
```
:::
::: {.fragment .current-visible fragment-index="0"}
```{r}
#| label: fig-intcomp-hyper-1
#| echo: false
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-alt: "Hypergeometric sampling distribution with one hundred sample means; every sample mean carries its interval: exact in the first state, Chebyshev in the second, bootstrap in the third, with one sample highlighted. Green line at the population proportion. State 2 of 3."
intcomp.plot("hypergeometric", 2)
```
:::
::: {.fragment fragment-index="1"}
```{r}
#| label: fig-intcomp-hyper-2
#| echo: false
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-alt: "Hypergeometric sampling distribution with one hundred sample means; every sample mean carries its interval: exact in the first state, Chebyshev in the second, bootstrap in the third, with one sample highlighted. Green line at the population proportion. State 3 of 3."
intcomp.plot("hypergeometric", 3)
```
:::
::::
::::
:::::

We now have three ways to calibrate an interval estimate.

1. [**Exact.**]{.pink}
   - Needs the whole sampling distribution.
   - We have it for binary responses, sampled with or without replacement.
   - Outside that, a pain.
2. [**Chebyshev.**]{.teal}
   - Needs only the standard deviation.
   - Conservative. Who wants to deny themselves precision that's really there?
3. **The bootstrap.**
   - Works, often.
   - Not when sampling without replacement, unless $n$ is tiny relative to $m$.
   - A thing you do, not a thing that helps you understand.

## The Bootstrap

```{r}
#| cache: false 
set.seed(default.seed)
samples = 1:10000 |> map(\(.) {
    J = sample(1:m, n, replace=TRUE)
    X = x[J]
    Y = y[J]
    p0 = mean(Y[X==0])
    p1 = mean(Y[X==1])
    data.frame(turnout=mean(Y), 
               difference=mean(Y[X==1]) - mean(Y[X==0]), 
               difference.sd=sqrt(p1*(1-p1)/sum(X==1) + p0*(1-p0)/sum(X==0)))
}) |> bind_rows() 

bootstrap.samples = 1:10000 |> map(\(.) { 
    I = sample(1:n, n, replace=TRUE)
    Xstar = X[I]
    Ystar = Y[I]
    data.frame(turnout=mean(Ystar), 
               difference=mean(Ystar[Xstar==1]) - mean(Ystar[Xstar==0]))
}) |> bind_rows() 
```
```{r}
#| cache: false

m.small = 1200
theta.small = mean(y[1:m.small])

Jsmall = sample(1:m.small, n, replace=FALSE)
Ysmall = y[Jsmall]
Xsmall = x[Jsmall]

samples.half = 1:10000 |> map(\(.) { 
    Jsmall = sample(1:m.small, n, replace=FALSE)
    Ysmall = y[Jsmall]
    Xsmall = x[Jsmall]
    data.frame(turnout=mean(Ysmall),
               difference=mean(Ysmall[Xsmall==1]) - mean(Ysmall[Xsmall==0]))
}) |> bind_rows() 

bootstrap.samples.half = 1:10000 |> map(\(.) { 
  I = sample(1:n, n, replace=TRUE)
  Ystar = Ysmall[I]
  Xstar = Xsmall[I]
  data.frame(turnout=mean(Ystar),
             difference=mean(Ystar[Xstar==1]) - mean(Ystar[Xstar==0]))
}) |> bind_rows() 
```
```{r}
#| label: bootstrap-dots-setup
#| cache: false
# Bootstrap slide figure. The actual sampling distribution of the turnout proportion with the
# dot cloud; in state k, dot k is highlighted, its bootstrap sampling distribution is drawn in
# its color with the middle 95% shaded, and the interval of that width sits on the dot.
set.seed(default.seed)
boot.J = list(sample(1:m, n, replace=TRUE), sample(1:m, n, replace=TRUE), sample(1:m, n, replace=TRUE))
stopifnot(all(sapply(boot.J, \(J) mean(y[J])) == samples$turnout[1:3]))
boot.dists = lapply(boot.J, \(J) { Yk = y[J]; map_vec(1:4000, \(.) mean(Yk[sample(1:n, n, replace=TRUE)])) })
boot.height = max(hist(samples$turnout, breaks=100, plot=FALSE)$density)
boot.xlim = range(samples$turnout)
boot.heights = annotation_dot_heights(100, 3, boot.height, default.seed)
boot.plot = function(k) {
  q = quantile(boot.dists[[k]], c(.025, .975))
  color = default.highlight.colors[k]
  ggplot() + 
    geom_histogram(aes(x=samples$turnout, y=after_stat(density)), bins=100, alpha=.1, linewidth=.1, fill='black', color='black') +
    with_annotations(samples$turnout, estimand=mean(y), height=boot.height, dots=100, highlight.dots=0, vline.alpha=NA, seed=default.seed) +
    geom_histogram(aes(x=boot.dists[[k]], y=after_stat(density)), bins=100, alpha=.2, linewidth=.1, fill=color, color=color) +
    annotate("rect", xmin=q[1], xmax=q[2], ymin=0, ymax=Inf, fill=color, alpha=.15) +
    annotate("pointrange", x=samples$turnout[k], y=boot.heights[k], xmin=q[1] - mean(boot.dists[[k]]) + samples$turnout[k], xmax=q[2] - mean(boot.dists[[k]]) + samples$turnout[k], color=color) +
    coord_cartesian(xlim=boot.xlim) + scale_y_continuous(breaks=NULL) + labs(x=NULL, y=NULL) + simplified.theme
}
```

:::: {.r-stack}
::: {.fragment .fade-out fragment-index="0"}
```{r}
#| label: fig-bootstrap-dot-1
#| echo: false
#| fig-width: 11
#| fig-height: 3.4
#| fig-alt: "Sampling distribution of the turnout proportion in gray with one hundred sample means floating over it. Sample 1 is highlighted in its color with its bootstrap sampling distribution drawn over the actual one, the middle 95 percent of it shaded, and an interval of that width on the dot."
boot.plot(1)
```
:::
::: {.fragment .current-visible fragment-index="0"}
```{r}
#| label: fig-bootstrap-dot-2
#| echo: false
#| fig-width: 11
#| fig-height: 3.4
#| fig-alt: "Sampling distribution of the turnout proportion in gray with one hundred sample means floating over it. Sample 2 is highlighted in its color with its bootstrap sampling distribution drawn over the actual one, the middle 95 percent of it shaded, and an interval of that width on the dot."
boot.plot(2)
```
:::
::: {.fragment fragment-index="1"}
```{r}
#| label: fig-bootstrap-dot-3
#| echo: false
#| fig-width: 11
#| fig-height: 3.4
#| fig-alt: "Sampling distribution of the turnout proportion in gray with one hundred sample means floating over it. Sample 3 is highlighted in its color with its bootstrap sampling distribution drawn over the actual one, the middle 95 percent of it shaded, and an interval of that width on the dot."
boot.plot(3)
```
:::
::::

- The bootstrap. A general method for estimating sampling distributions.
  1. Draw a sample of size $n$ from your sample. *A bootstrap sample.*
  2. Calculate your estimator on it. *A bootstrap estimate.*
  3. Repeat. *The bootstrap sampling distribution.*
- This simulates sampling with replacement.
  - Not sampling without. There's no natural analog.
- Great for what we *have learned* after we have data.
  - Not for what we *can learn* before we have it.
- Today: a tool for that. *Normal approximation.*
  - And a job for it: *sample size calculation*. Choosing $n$ so your interval is as narrow as you want.


# Normal Approximation

## The Normal Distribution

::: zoom-free
```{r}
#| layout-ncol: 2
#| fig-width: 7

expand.grid(x=seq(-5,5,by=.1), theta=c(-2,0,2)) |>
  ggplot() + 
  geom_area(aes(x=x, y=dnorm(x, theta, 1), fill=factor(theta)), position='identity', alpha=.3) +
  guides(fill=guide_legend(title='θ'), color='none')

expand.grid(x=seq(-5,5,by=.1), sigma=c(.5,1,2)) |>
  ggplot() + 
  geom_area(aes(x=x, y=dnorm(x, 0, sigma), fill=factor(sigma)), position='identity', alpha=.3) +
  guides(fill=guide_legend(title='σ'), color='none')
```
:::

- The normal distribution is a parametric family of distributions involving two parameters.
  1. Its mean, $\theta$. 
  2. Its standard deviation, $\sigma$.
- We say a random variable $X$ is normally distributed with mean $\theta$ and standard deviation $\sigma$ ...
- ... if the probability that it's in an interval $[a,b]$ is given by this integral of its *probability density*.
$$
P_{\theta,\sigma}(X \in [a,b]) = \int_a^b  f_{\theta,\sigma}(x)dx \qfor f_{\theta, \sigma}(x) = \frac{1}{\sqrt{2\pi}\sigma}e^{-\frac{(x-\theta)^2}{2\sigma^2}}
$$


- We have to talk about the probability that it's 'in an interval' rather than that it 'takes a value' because ...
- ... the probability it actually takes on any particular value is $x$ zero: it's an integral from $x$ to $x$. 

## Normal Approximation
```{r}
p.even = ggplot() + 
  geom_col(aes(x=(0:10)/10,  y=10*dbinom(0:10,  10,  .5)), alpha=.2, linewidth=.1,       fill=pollb, color=pollb) +
  geom_col(aes(x=(0:30)/30,  y=30*dbinom(0:30, 30, .5)), alpha=.2, linewidth=.1, fill=pollc, color=pollc) + 
  geom_col(aes(x=(0:90)/90,  y=90*dbinom(0:90, 90, .5)), alpha=.2, linewidth=.1, fill=polld, color=polld) +
           labs(x='', y='')

p.uneven = ggplot() + 
  geom_col(aes(x=(0:10)/10,    y=10*dbinom(0:10,  10,  .5)), alpha=.2, linewidth=.1, fill=pollb, color=pollb) +
  geom_col(aes(x=(0:25)/25,    y=25*dbinom(0:25, 25, .5)), alpha=.2, linewidth=.1, fill=pollc, color=pollc) + 
  geom_col(aes(x=(0:83)/83,    y=83*dbinom(0:83, 83,  .5)), alpha=.2, linewidth=.1, fill=polld, color=polld)

xx = seq(0,1,by=.01)

p.even.plus.normal = p.even +
  geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/10))), color=pollb) + 
  geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/30))), color=pollc) + 
  geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/90))), color=polld)

p.uneven.plus.normal = p.uneven + 
  geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/10))), color=pollb) + 
  geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/25))), color=pollc) + 
  geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/83))), color=polld)
```

```{r}
#| fig-width: 4
#| layout-ncol: 3
 
pbinom.10 = ggplot() + geom_col(aes(x=(0:10)/10,  y=10*dbinom(0:10,  10,  .5)), 
                    alpha=.2, linewidth=.1, fill=pollb, color=pollb) +
            geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/10))), color=pollb)

pbinom.30 = ggplot() + geom_col(aes(x=(0:30)/30,  y=30*dbinom(0:30, 30, .5)), 
                    alpha=.2, linewidth=.1, fill=pollc, color=pollc) +
            geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/30))), color=pollc)

pbinom.90 = ggplot() + geom_col(aes(x=(0:90)/90,  y=90*dbinom(0:90, 90, .5)), 
                    alpha=.2, linewidth=.1, fill=polld, color=polld) +
            geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt(.5*.5/90))), color=polld)

scale_y = scale_y_continuous(breaks=seq(0,8,by=1), limits=c(0,8))
scale_x = scale_x_continuous(breaks=seq(0,1,by=.25), limits=c(0,1))
pbinom.10 + scale_x + scale_y
pbinom.30 + scale_x + scale_y
pbinom.90 + scale_x + scale_y
```

- A distribution's normal approximation is a normal distribution with the same mean $\theta$ and standard deviation $\sigma$.
  - These show up everywhere because they're easy to work with and the approximation tends to be good.
  - Approximation is good, in particular, for the distribution of a mean of independent random variables.
  - Or independent enough ones.
- Above, we see three Binomial distributions with their normal approximations:
  - the distributions of the mean of [10]{.pollb}, [30]{.pollc}, and [90]{.polld} coin flips.
  - These approximations get increasingly accurate as the number of flips $n$ increases. 
  - That's universal. It always happens with means. It's called the *Central Limit Theorem*.

::: {.callout-note}
### The Central Limit Theorem
The sampling distribution of a mean of independent random variables, $\faded \frac{1}{n}\sum\limits_{i=1}^n X_i$,
gets arbitrarily close to its normal approximation \
as the number of random variables $\faded n$ increases to infinity. 

*Caveats*.

- Finite variance. Without it, there is no normal limit.
- No single term dominates. Each one's share of the mean's variance has to be infinitesimal as $\faded n \to \infty$.
- Independence, or close to it. Dependence changes the variance, and enough of it changes the shape.
:::

## The Width of Normal Distributions 

```{r}
#| out-height: 300px
#| fig-asp: .3 

breaks = seq(-4, 4, 1) 
signs = case_when(
  breaks < 0 ~ '-', 
  breaks == 0 ~ '',
  breaks > 0 ~ '+')
offsets = case_when(
  breaks != 0 ~ ifelse(abs(breaks)==1, 'σ', sprintf('%dσ', abs(breaks))),
  breaks == 0 ~ '')
labels.abstract = sprintf('θ %s %s', signs, offsets)

x=seq(-4,4,by=.01)
normal.dist.plot = ggplot() + 
       geom_area(aes(x=x,y=dnorm(x)), alpha=.1, color='black', fill='black') + 
       geom_vline(aes(xintercept=qnorm(.5)), linetype='solid', color='blue') +
       geom_vline(aes(xintercept=qnorm(c(.16,.84))), linetype='dashed', color='blue') +
       geom_vline(aes(xintercept=qnorm(c(.025,.975))), linetype='dotted', color='blue') +
       scale_x_continuous(breaks=seq(-4,4,by=1), labels=labels.abstract) 

normal.dist.plot
```

- One thing that's convenient about the normal distribution is that it's easy to reason about.
- In particular, it's easy to reason about the width of its middle *x*%.
  - To include `r 100*round(pnorm(1)-pnorm(-1),3)`% of draws, you go out 1 standard deviation from its mean.
  - To include `r 100*round(pnorm(2)-pnorm(-2),3)`% of draws, you go out 2 standard deviations.
  - To include `r 100*round(pnorm(3)-pnorm(-3),3)`% of draws, you go out 3 standard deviations.
- To get almost exactly 95% of draws, we go out 1.96 standard deviations: $\theta \pm 1.96\sigma$.
  - Chebyshev got us 95% at $\sqrt{20} \approx 4.5$ standard deviations. The normal gets it at $1.96$.
  - Same standard deviation. An interval $2.3\times$ narrower.
- Two is close enough in practice, but we tend to write 1.96 anyway.
  - It's a signal about what we're doing. You can get a 2 anywhere in a calculation. 
  - When you see a 1.96, you know you're talking about the middle 95% of the normal.

## Calibration Using Normal Approximation

```{r}
#| out-height: 300px
#| fig-asp: .3 

theta = .5
sigma = sqrt(theta*(1-theta)/90)  
binom.90.draws = rbinom(1000, 90, .5)/90
theta.hat = binom.90.draws[1]
sigma.hat = sqrt(theta.hat*(1-theta.hat)/90)  
wbinom = width_binom(theta.hat, 90)
wnorm = 2*1.96*sigma.hat 

breaks = seq(-4, 4, 1) 
signs = case_when(
  breaks < 0 ~ '-', 
  breaks == 0 ~ '',
  breaks > 0 ~ '+')
offsets = case_when(
  breaks != 0 ~ ifelse(abs(breaks)==1, 'σ', sprintf('%dσ', abs(breaks))),
  breaks == 0 ~ '')
labels.abstract = sprintf('θ %s %s', signs, offsets)

x=seq(theta-4*sigma, theta+4*sigma,by=.01)
proportion.normal.dist.plot = ggplot() + 
       geom_area(aes(x=x,y=dnorm(x, theta.hat, sigma.hat)), alpha=.1, color='blue', fill='blue', linewidth=.1) + 
       geom_vline(aes(xintercept=qnorm(.5, theta.hat, sigma.hat)), linetype='solid', color='blue') +
       geom_vline(aes(xintercept=qnorm(c(.16,.84), theta.hat, sigma.hat)), linetype='dashed', color='blue') +
       geom_vline(aes(xintercept=qnorm(c(.025,.975), theta.hat, sigma.hat)), linetype='dotted', color='blue') +
       scale_x_continuous(breaks=seq(theta.hat-4*sigma.hat,theta.hat+4*sigma.hat, by=sigma.hat), labels=labels.abstract, limits=c(theta-4*sigma, theta+4*sigma)) 
```

```{r}
#| fig-width: 7
#| layout-ncol: 2
#| fig-cap: 
#|   - The (Binomial) sampling distribution and its normal approximation.
#|   - Binomial and normal estimates of the sampling distribution.

actual.sampling.dist.plot = pbinom.90 + 
  with_annotations(binom.90.draws,  estimand=theta, highlight.dots=0, height=8) +
  geom_pointrange(aes(y=3, x=theta.hat, xmin=theta.hat-wbinom/2, xmax=theta.hat+wbinom/2), color='red') +
  geom_pointrange(aes(y=3.2, x=theta.hat, xmin=theta.hat-wnorm/2,  xmax=theta.hat+wnorm/2), color='blue') +
  scale_x_continuous(limits=c(theta-4*sigma, theta+4*sigma))

xdiscrete = seq(0,1,by=1/90)
estimated.sampling.dist.plot = proportion.normal.dist.plot + 
  geom_vline(xintercept=theta, color='green', linewidth=3, alpha=.8) +
   geom_col(aes(x=xdiscrete, y=90*dbinom(90*xdiscrete, 90, theta.hat)), 
            alpha=.1, color='red', fill='red', linewidth=.1) +
  scale_x_continuous(limits=c(theta-4*sigma, theta+4*sigma)) + 
  geom_pointrange(aes(y=3, x=theta.hat, xmin=theta.hat-wbinom/2, xmax=theta.hat+wbinom/2), color='red') +
  geom_pointrange(aes(y=3.2, x=theta.hat, xmin=theta.hat-wnorm/2,  xmax=theta.hat+wnorm/2), color='blue') 

actual.sampling.dist.plot
estimated.sampling.dist.plot
```

- Calibrating interval estimates using normal approximation is easy.
  - You estimate the standard deviation of your point estimator.
  - You go out &plusmn; 1.96 (estimated) standard deviations from your point estimate.

$$
\text{interval estimate} = \hat\theta \pm 1.96 \hat\sigma
$$

- We're choosing our interval's width essentially the same way we always have.
  - We're still using the middle 95% of an estimate of the sample distribution.
  - It just looks different because we have a convenient formula for that width.
- Above left, you can see two interval estimates superimposed on the sampling distribution of a sample proportion.
  - The [red one]{.red} is calibrated using the binomial as before.
  - The [blue one]{.blue} is calibrated using normal approximation.
- Above right, you can see the [binomial]{.red} and [normal]{.blue} sampling distribution estimates these are based on.

## When This Works

```{r}
#| fig-width: 7
#| layout-ncol: 2
#| fig-cap: 
#|   - The (Binomial) sampling distribution and its normal approximation.
#|   - Binomial and normal estimates of the sampling distribution.

actual.sampling.dist.plot
estimated.sampling.dist.plot
```

::: {style="text-align: center"}
This works if three things are true.
:::

1. The sampling distribution needs to be in the right place, i.e. centered on the estimation target.
   - This is called *unbiasedness* of our point estimator.
   - Most of the estimators we'll talk about in this class are unbiased or almost unbiased.
2. The sampling distribution needs to be approximately normal.
   - The CLT tells us this tends to happen, especially for large $n$.
   - We'll talk about how accurate it is next week. For most of the estimators we'll study, you can take it as a given.
3. The estimated sampling distribution has approximately the right width.
   - This amounts to getting a good estimate of our point estimator's standard deviation.
   - We'll work on this, again for the sample proportion, later today. And for other estimators throughout the semester.
   - It is, however, a bit of a pain. That's one reason you might prefer the bootstrap.


# Normal Approximation in Context

## A Proportion when Sampling with Replacement

```{r}
#| fig-width: 4
#| layout-ncol: 3 

scale_y = scale_y_continuous(breaks=seq(0,12,by=1), limits=c(0,12))

pbinom.10 + scale_x + scale_y
pbinom.30 + scale_x + scale_y
pbinom.90 + scale_x + scale_y
```

- When we sample from a binary population with replacement, our sample proportion's distribution is Binomial.
  - And we can estimate that distribution by plugging our sample proportion into the Binomial formula.
  - But its normal approximation tends to be very good, so we can get away with estimating that instead.
  - Its standard deviation is the one we worked out last class.

$$
\begin{aligned}
\text{normal approximation}     & \ f_{\theta,\sigma}(x) \qfor  \sigma^2 = \frac{\theta(1-\theta)}{n} \\
\text{corresponding estimate}   & \ f_{\hat\theta, \hat \sigma}(x) \qfor \hat\sigma^2 = \frac{\hat \theta(1-\hat \theta)}{n}. 
\end{aligned}
$$

- Pictured: three Binomial distributions with their normal approximations. 
  - These are the distributions of the sample proportion when we draw a samples of size [10]{.pollb}, [30]{.pollc}, and [90]{.polld} ...
  - ... with replacement from a binary population in which the proportion of ones is $\theta=.5$.

## A Proportion when Sampling without Replacement

```{r}
#| fig-width: 4
#| layout-ncol: 3 


phyper.10 = ggplot() + geom_col(aes(x=(0:10)/10,  y=10*dhyper(0:10, 10, 10, 10)), 
                    alpha=.2, linewidth=.1, fill=pollb, color=pollb) +
            geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt((.5*.5/10) * (10-1)/(2*10-1)))), color=pollb)

phyper.30 = ggplot() + geom_col(aes(x=(0:30)/30,  y=30*dhyper(0:30, 30, 30, 30)), 
                    alpha=.2, linewidth=.1, fill=pollc, color=pollc) +
            geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt((.5*.5/30) * (30-1)/(2*30 - 1)))), color=pollc)

phyper.90 = ggplot() + geom_col(aes(x=(0:90)/90,  y=90*dhyper(0:90, 90, 90, 90)), 
                    alpha=.2, linewidth=.1, fill=polld, color=polld) +
            geom_line(aes(x=xx, y=dnorm(xx, .5, sqrt((.5*.5/90) * (90-1)/(2*90 - 1)))), color=polld)

phyper.10 + scale_x + scale_y
phyper.30 + scale_x + scale_y
phyper.90 + scale_x + scale_y
```

- Sampling *without* replacement from a binary population, our sample proportion is Hypergeometric.
  - We can estimate it by plugging our sample proportion into the Hypergeometric formula.
  - But again, its normal approximation is very good, so we can estimate that instead.
- This shows where we go wrong using the Binomial---or equivalently the bootstrap.
  - When $\faded n$ is a meaningful fraction of $\faded m$, the Binomial is too wide.
  - Its standard deviation is the Binomial's times $\faded \sqrt{\frac{m-n}{m-1}}$. A homework problem.
  - When we're sampling half our population, i.e. $n=m/2$, that's roughly $\faded \sqrt{1/2} \approx .7$. 
  

$$
\begin{aligned}
\text{normal approximation}     & \ f_{\theta,\sigma}(x) \qfor \sigma^2 = \frac{\theta(1-\theta)}{n} \times  \frac{m-n}{m-1} \\
\text{corresponding estimate}   & \ f_{\hat\theta, \hat \sigma}(x) \qfor \hat\sigma^2 = \frac{\hat \theta(1-\hat \theta)}{n} \times \frac{m-n}{m-1}. 
\end{aligned}
$$

- Pictured: three Hypergeometric distributions with their normal approximations. 
  - Sample proportions for samples of size [10]{.pollb}, [30]{.pollc}, and [90]{.polld} ...
  - ... drawn without replacement from a binary population twice the size with $\faded \theta=.5$.

## A Mean when Sampling with Replacement

```{r}
#| label: vote-history-setup
# Vote history: number of elections voted in out of the last 5, as in the bootstrap chapter.
set.seed(42)
h = rep(NA, m)
h[y == 1] = rbinom(sum(y == 1), size = 5, prob = 0.7)
h[y == 0] = rbinom(sum(y == 0), size = 5, prob = 0.3)
H = h[J1]
H.bar = mean(H)
sigma.hat.H = sd(H)/sqrt(n)
```

```{r}
#| label: fig-vote-history-normal
#| fig-width: 7
#| fig-height: 3.8
#| layout-ncol: 2
#| fig-alt:
#|   - "Histogram of the 625 sampled voters' vote histories, 0 through 5 elections, with a point at the sample mean and arms one sample standard deviation to either side."
#|   - "Histogram of ten thousand sample means of vote history drawn from the population, with the normal approximation drawn over it in blue."
ggplot() + 
  geom_bar(aes(x=H, y=after_stat(prop)), alpha=.3) +
  geom_pointrange(aes(x=H.bar, xmin=H.bar-sd(H), xmax=H.bar+sd(H), y=max(prop.table(table(H)))/2), color='blue') +
  labs(x='elections voted in, of the last 5', y='')

mean.history.samples = map_vec(1:10000, \(.) mean(h[sample(1:m, n, replace=TRUE)]))
xx.h = seq(min(mean.history.samples), max(mean.history.samples), length.out=500)
ggplot() + 
  geom_histogram(aes(x=mean.history.samples, y=after_stat(density)), bins=100, alpha=.3) +
  geom_area(aes(x=xx.h, y=dnorm(xx.h, mean(h), sd(h)/sqrt(n))), color='blue', fill='blue', alpha=.3, linewidth=.1) +
  labs(x='sample mean vote history', y='')
```

- Not special to proportions. Take the average vote history $\bar H_n$.
- Its standard deviation is the population standard deviation over $\sqrt{n}$. Plug in the sample's.

$$
\begin{aligned}
\text{normal approximation}     & \ f_{\mu,\sigma_n}(x) \qfor \sigma_n^2 = \frac{\sigma_H^2}{n} \qqtext{for} \sigma_H^2 = \frac{1}{m}\sum_{j=1}^m (h_j - \mu)^2 \\
\text{corresponding estimate}   & \ f_{\bar H_n, \hat \sigma_n}(x) \qfor \hat\sigma_n^2 = \frac{\hat\sigma_H^2}{n} \qqtext{for} \hat\sigma_H^2 = \frac{1}{n}\sum_{i=1}^n (H_i - \bar H_n)^2.
\end{aligned}
$$

- Pictured: left, the `r n` sampled voters' vote histories with $\bar H_n$ and the sample standard deviation.
  - Right, the sampling distribution of $\bar H_n$ over draws from the population, with its normal approximation.

## Using Normal Approximation 

```{r}
#| label: fig-using-normal-approx-strip
#| fig-width: 4
#| layout-ncol: 3
#| fig-alt:
#| - "Binomial distribution and normal approximation for the sample proportion from 10 fair draws."
#| - "Binomial distribution and normal approximation for the sample proportion from 30 fair draws."
#| - "Binomial distribution and normal approximation for the sample proportion from 90 fair draws."
pbinom.10 + scale_x + scale_y
pbinom.30 + scale_x + scale_y
pbinom.90 + scale_x + scale_y
```

\begin{aligned}
&f_{\hat\theta, \hat \sigma}(x) \qfor \hat\sigma^2 = \frac{\hat \theta(1-\hat \theta)}{n} && \text{a proportion when sampling with replacement} \\
&f_{\hat\theta, \hat \sigma}(x) \qfor \hat\sigma^2 = \frac{\hat \theta(1-\hat \theta)}{n} \times \frac{m-n}{m-1} && \text{a proportion when sampling without replacement} \\
&f_{\bar H_n, \hat \sigma_n}(x) \qfor \hat\sigma_n^2 = \frac{1}{n} \times \frac{1}{n}\sum_{i=1}^n (H_i - \bar H_n)^2 && \text{a mean when sampling with replacement}
\end{aligned}


- What makes all this work is that we're using good estimates of our estimator's standard deviation.
  - If we have a formula for the estimator's standard deviation, this is usuallly not so hard.
  - We estimate whatever population summaries show up in the formula and plug them in.


# Sample Size Calculation

## An Interval for Turnout in 2020

```{r}
#| out-height: 300px 
#| fig-asp: .3
 
theta = mean(y)
theta.hat = mean(Y)
hbinom = n*max(dbinom(0:n, n, theta.hat))
wbinom = width_binom(theta.hat, n)
wnorm = 2*1.96*sqrt(theta.hat*(1-theta.hat)/n)
w95 = wbinom

xx = seq(.6,.8, length=1000)
turnout.histogram = ggplot() + 
  geom_col(aes(x=(0:n)/n, y=n*dbinom(0:n, n, theta.hat)), alpha=.3, color='red', fill='red',linewidth=.1) + 
  geom_area(aes(x=xx, y=dnorm(xx, theta.hat, sqrt(theta.hat*(1-theta.hat)/n))), color='blue', fill='blue', alpha=.3, linewidth=.1) + 
  geom_pointrange(aes(y=.3*hbinom, x=theta.hat, xmin=theta.hat-wbinom/2, xmax=theta.hat+wbinom/2), color='red') +
  geom_pointrange(aes(y=.34*hbinom, x=theta.hat, xmin=theta.hat-wnorm/2,  xmax=theta.hat+wnorm/2), color='blue') 
turnout.histogram + scale_x_continuous(limits=c(.6, .8))
```

- Let's compare our two approaches to calibration in our turnout poll.

$$
\begin{aligned}
\textcolor{red}{\text{binomial interval}} &= `r sprintf("%0.4f", round(theta.hat,4))`  \pm `r sprintf("%0.4f", round(wbinom/2,4))`  \\
\textcolor{blue}{\text{normal interval}}    &= `r sprintf("%0.4f", round(theta.hat,4))`  \pm `r sprintf("%0.4f", round(wnorm/2,4))` 
\end{aligned}
$$

- We have to go out to 4 digits, way beyond what's statistically meaningful, to see any difference in these intervals.
  - Our interval estimate---either one---is telling you might be off by a few hundredths. 
  - Who cares about another couple ten-thousandths at the edge of the interval?

## Thinking Speculatively About Intervals


```{r}
#| out-height: 300px 
#| fig-asp: .3
 
turnout.histogram + scale_x_continuous(limits=c(.6, .8))  
```

- Suppose we're not satisfied with this level of precision, so we're going to collect more data.
- Suppose we want our interval to be $\pm .01$ instead of $\pm `r round(w95/2, 3)`$. 
- How many people, in total, do we need to call?  We can use the normal approximation to figure that out.

$$
\hat \theta \pm 1.96\sigma_n \qfor \sigma_n = \sqrt{\frac{\theta (1-\theta)}{n}} \qqtext{ is } \hat\theta \pm .01 \qqtext{if} 1.96\sqrt{\frac{\theta (1-\theta)}{n}} = .01
$$

- Now all we have to do is solve for $n$. And, since we don't know $\theta$, use our best guess, $\hat\theta=`r mean(Y)`$.

::: fragment

$$ 
n = \frac{1.96^2 \ \hat\theta (1-\hat\theta)}{.01^2} \approx `r round(1.96^2*mean(Y)*(1-mean(Y))/.01^2,-3)`
$$

:::

## The Easy Version

- There's a trick to this. Let's compare the interval width we have to the one we want.

$$
\begin{aligned}
\pm `r round(w95/2,3)` &= \pm 1.96\hat\sigma_{625} \qfor \hat\sigma_{625} = \sqrt{\frac{\hat\theta (1-\hat\theta)}{625}}  \\
\pm 0.01 &= \pm 1.96\hat\sigma_{n} \qfor \hat\sigma_{n} = \sqrt{\frac{\hat\theta (1-\hat\theta)}{n}}   
\end{aligned}
$$

- All that changes in this formula is the sample size. And the *sample size ratio* falls out of the interval width ratio.

$$
\frac{`r round(w95/2,3)`}{.01} 
= \frac{1.96\hat\sigma_{625}}{1.96\hat\sigma_n}
= \frac{\sqrt{\frac{\hat\theta (1-\hat\theta)}{625}}}{\sqrt{\frac{\hat\theta (1-\hat\theta)}{n}}}
 = \sqrt{\frac{n}{625}}
\qqtext{ so } n = 625\left(\frac{`r round(wnorm/2,3)`}{.01}\right)^2 \approx `r round(625*((wnorm/2)/.01)^2,-3)`
$$

- To get the new sample size, multiply ...
  - the sample size we have
  - the square of the desired ratio of the interval widths.
- To double precision, quadruple the sample size. To triple it, increase it 9x.
- To get another digit, i.e. increase precision 10x, increase the sample size 100x.

## Starting from Scratch

```{r}
#| label: fig-binomials-across-thetas
#| out-height: 300px
#| fig-asp: .3
#| fig-alt: "Nine binomial sampling distributions of the sample proportion for population proportions from 0.1 through 0.9 at n = 625, each with its calibrated 95 percent interval drawn as a horizontal belt; the belt is widest at one half."
thetas = seq(.1, .9, by=.1)
pmfs = expand_grid(theta = thetas, x = (0:n)/n) |>
  mutate(p = dbinom(x*n, n, theta), y=0, width=0)

for(theta in thetas) {
  w = width_binom(theta, n, alpha=.05)
  thispmf = pmfs[pmfs$theta==theta,] 
  belt.y = thispmf$p[which.min(abs(thispmf$x - (theta-.5*w)))] 
  pmfs$y[pmfs$theta==theta] = belt.y
  pmfs$width[pmfs$theta==theta] = w
}

binomials.across.thetas = ggplot(pmfs) +
    geom_area(aes(x=x, y=p, fill=factor(theta), color=factor(theta)), position='identity', alpha=.1) +
    geom_linerange(aes(xmin=theta-.5*width, xmax=theta+.5*width, y=y, color=factor(theta))) +
    scale_x_continuous(breaks=seq(0,1,by=.1), limits=c(0,1)) +
    scale_fill_discrete(name="theta") + guides(color='none') 
binomials.across.thetas
```

- What do we do if we don't have any data yet? 
  - Then we don't have a 'current interval width' to compare to the 'desired interval width'.
  - And we don't have a sample proportion $\hat\theta$ to plug in for the population proportion $\theta$.
- But we can still use the formula we worked out earlier to get somewhere.
$$
1.96\sqrt{\frac{\theta (1-\theta)}{n}} = .01 \qqtext{so} n = \frac{1.96^2 \ \theta (1-\theta)}{.01^2}
$$

- We don't have an estimate of $\theta$, but we do know it's between 0 and 1. And, consequently, so is $\theta(1-\theta)$.
- So we know that if we just substitute $1$ into our formula, we'll get a number that's bigger than we need.

$$
n < n' = \frac{1.96^2 \cdot 1}{.01^2} \approx `r sprintf("%d", round(1.96^2/.01^2,-2))`
$$

- That's a bit excessive. In fact, we can substitute in $1/4$ instead of $1$. 
$$
n < n' = \frac{1.96^2 \cdot 1/4}{.01^2} \approx `r sprintf("%d", round(1.96^2/4/.01^2,-2))`
$$

- Much better. That's pretty close to the number we got *with* preliminary data. 

## Why Can We Use 1/4?
```{r}
#| label: fig-why-a-quarter
#| layout-ncol: 2
#| fig-width: 7
#| fig-alt:
#|   - "Nine binomial sampling distributions for population proportions from 0.1 through 0.9 with their 95 percent belts, widest at one half."
#|   - "Curve of the Bernoulli variance theta times one minus theta from zero to one; largest at theta equal to one half."
binomials.across.thetas
xx = seq(0,1,by=.001)
ggplot() + geom_line(aes(x=xx, y=xx*(1-xx)))
```

- Here's the claim. Why is it true? 
$$ 
n := \frac{1.96^2 \ \theta (1-\theta)}{.01^2} < n' := \frac{1.96^2 \ \times 1/4}{.01^2}
$$

- **Hint.** Look at the graph of $f(x) = x(1-x)$ above.


::: fragment

 - Because $1/4$ is the biggest $\theta (1-\theta)$ gets for $\theta \in [0,1]$.
 - And it happens, for what it's worth, when $\theta = 1/2$. 
 - That is, we'll have the least precision --- at a given sample size --- when the proportion we're estimating is 1/2. 

:::


# Case Study  
The NSW Experiment

## The National Supported Work Demonstration

```{r}
#| label: nsw-load
nsw.data = read.csv('data/nsw.csv')
nsw = nsw.data |> filter(dataset=='nsw', treated)
nsw.Y = as.numeric(nsw$income78)
nsw.n = length(nsw.Y)
mu.hat = mean(nsw.Y)
sd.hat = sd(nsw.Y)
sigma.hat = sd.hat / sqrt(nsw.n)
# 1978 dollars to today's: CPI-U averaged 65.2 in 1978 and about 330 in 2026, a factor of 5.
nsw.inflation = 5
```
```{r}
#| label: nsw-bootstrap
nsw.bootstrap.samples = map_vec(1:10000, function(.) {
    I = sample(1:nsw.n, nsw.n, replace=TRUE)
    mean(nsw.Y[I])
})
```
```{r}
#| label: nsw-width
nsw.w95.boot = width(nsw.bootstrap.samples, alpha=.05)
nsw.interval = mu.hat + c(-1,1)*nsw.w95.boot/2
```
```{r}
#| label: nsw-interval-plot
nsw.interval.plot = ggplot() +
  geom_histogram(aes(x=nsw.bootstrap.samples), alpha=.3, bins=100) +
  geom_vline(xintercept=mu.hat, color='blue', alpha=.4) +
  geom_pointrange(aes(x=mu.hat, xmin=nsw.interval[1], xmax=nsw.interval[2], y=150), color='blue') +
  scale_x_continuous(labels=dollar) + labs(x='bootstrap sample means', y='')
```


:::: columns 
::: column
```{r}
#| label: fig-nsw-bootstrap
#| fig-width: 7
#| fig-height: 3.4
#| fig-alt: "Histogram of ten thousand bootstrap sample means of treated 1978 income, with a blue point at the sample mean and a blue interval spanning the middle 95 percent."
nsw.interval.plot
``` 

::: 
::: column

$n$ |  $\hat \mu$ | sample SD
--- | --- | ---
`r nsw.n` | `r round(mu.hat, -2)` | `r round(sd.hat, -2)`

:::
::::

::: {.panel-tabset}
### The Program

- The National Supported Work Demonstration was an experimental program in the 1970s that provided job training and other services.
- Here 'experimental' doesn't just mean untested. It was run as a big experiment.
- From a population of eligible people, they drew a sample of size `r sum(nsw.data$dataset=='nsw')` to participate.
- To help estimate the impact of the program, participants were randomized into two groups.
  - The 'treated' group received the training and services.
  - The 'control' group did not. 
- Today we'll look at the treated group only: `r nsw.n` people.

### The Result 

- A year or two after the program ended, they looked at the incomes of the people who'd been treated.
  - The average within the sample was about \$`r round(mu.hat, -2)` per year. In 1978 dollars.
  - A dollar in 1978 was worth about what \$`r nsw.inflation` is today.
  - So adjusting for inflation, that's an in-sample average of about \$`r sprintf("%d", round(mu.hat*nsw.inflation, -2))`.
- But the sample was relatively small. And, as a result, the interval estimate is pretty wide.

$$
\begin{aligned}
\mu 
&\in `r round(mu.hat,-2)` \pm `r round(nsw.w95.boot/2,-2)` \approx [`r round(nsw.interval[1], -2)`, \ `r round(nsw.interval[2], -2)`] 
&&\qqtext{is a 95\% confidence interval in 1978 dollars} \\
&\in `r sprintf("%d", round(mu.hat*nsw.inflation, -2))` \pm `r sprintf("%d", round(nsw.w95.boot*nsw.inflation/2, -2))` \approx [`r sprintf("%d", round(nsw.interval[1]*nsw.inflation, -2))`, \ `r sprintf("%d", round(nsw.interval[2]*nsw.inflation, -2))`] 
&&\qqtext{adjusted for inflation}
\end{aligned}
$$

:::


## Activities

:::: columns 
::: column
```{r}
#| label: fig-nsw-bootstrap-again
#| fig-width: 7
#| fig-height: 3.2
#| fig-alt: "The same bootstrap histogram of treated mean income with its blue 95 percent interval."
nsw.interval.plot
``` 

::: 
::: column

$n$ |  $\hat \mu$ | sample SD
--- | --- | ---
`r nsw.n` | `r round(mu.hat, -2)` | `r round(sd.hat, -2)`

:::
::::

::: {.panel-tabset}
### Warm-Up 

- Use the information in this table to calculate a 95% interval based on normal approximation.
- Compare it to the interval we got using the bootstrap sampling distribution for calibration.


### Resizing 
  
  - Suppose we want to know the population mean to within \$500 (1978 dollars).
  - How can we resize our study to make that happen?
  - We could run a second wave, drawing more people from the same population exactly as before, to get a larger sample like the one we have. 
  - What sample size would we need to be able to get a 95% confidence interval this accurate?

:::

## Solutions and Discussion

```{r}
#| label: nsw-normal-width
nsw.w95.norm = 2*1.96*sigma.hat
```

```{r}
#| label: fig-nsw-two-intervals
#| fig-width: 11
#| fig-height: 3.2
#| fig-alt: "The bootstrap histogram of treated mean income with the normal approximation drawn over it. Two interval bars: the blue bootstrap interval and the red normal-approximation interval, nearly the same length."
xx.nsw = seq(min(nsw.bootstrap.samples), max(nsw.bootstrap.samples), length.out=500)
nsw.interval.plot + 
  geom_line(aes(x=xx.nsw, y=10000*diff(range(nsw.bootstrap.samples))/100 * dnorm(xx.nsw, mu.hat, sigma.hat)), color='red') +
  geom_pointrange(aes(x=mu.hat, xmin=mu.hat-nsw.w95.norm/2, xmax=mu.hat+nsw.w95.norm/2, y=170), color='red')
```

::: {.fragment}
- **Warm-Up.** $\hat\mu \pm 1.96\,\hat\sigma_n$, with $\hat\sigma_n$ the sample standard deviation over $\sqrt{n}$.
  - $`r round(mu.hat, -2)` \pm 1.96 \times `r round(sd.hat, -2)`/\sqrt{`r nsw.n`} \approx `r round(mu.hat, -2)` \pm `r round(nsw.w95.norm/2, -2)`$. The bootstrap gave $\pm `r round(nsw.w95.boot/2, -2)`$.
:::

::: {.fragment}
- **Resizing, the hard way.** Just like the turnout poll: we want $\pm 500$ instead of $\pm `r round(nsw.w95.norm/2, -1)`$.

$$
\hat\mu \pm 1.96\hat\sigma_n \qfor \hat\sigma_n = \frac{\hat\sigma_Y}{\sqrt{n}} \qqtext{ is } \hat\mu \pm 500 \qqtext{if} 1.96\frac{\hat\sigma_Y}{\sqrt{n}} = 500
$$

- Now all we have to do is solve for $n$, using the sample standard deviation $\hat\sigma_Y = `r round(sd.hat, -2)`$.

$$
n = \frac{1.96^2 \ \hat\sigma_Y^2}{500^2} \approx `r round((1.96*sd.hat/500)^2, -1)`
$$
:::

::: {.fragment}
- **Resizing, the easy way.** Compare the interval width we have to the one we want.

$$
\begin{aligned}
\pm `r round(nsw.w95.norm/2, -1)` &= \pm 1.96\hat\sigma_{`r nsw.n`} \qfor \hat\sigma_{`r nsw.n`} = \frac{\hat\sigma_Y}{\sqrt{`r nsw.n`}} \\
\pm 500 &= \pm 1.96\hat\sigma_n \qfor \hat\sigma_n = \frac{\hat\sigma_Y}{\sqrt{n}}
\end{aligned}
$$

- All that changes is the sample size. The *sample size ratio* falls out of the interval width ratio.

$$
\frac{`r round(nsw.w95.norm/2, -1)`}{500} = \frac{1.96\hat\sigma_{`r nsw.n`}}{1.96\hat\sigma_n} = \sqrt{\frac{n}{`r nsw.n`}}
\qqtext{ so } n = `r nsw.n`\left(\frac{`r round(nsw.w95.norm/2, -1)`}{500}\right)^2 \approx `r round(nsw.n*((nsw.w95.norm/2)/500)^2, -1)`
$$
:::
